---
title: "Statistics: Final Report"
output: html_document
---

```{r bio239 line 6, include = FALSE, echo = FALSE}
require(tidyverse)
require(knitr)
```

<div style = "margin-bottom:50px;">

Download the package `tidyverse`

</div>

<div style = "margin-bottom:50px;">
```{r bio239 line 18, echo = FALSE}
data <- read.csv("bio239_spring2022.csv")
knitr::kable(head(data[, 1:5]), "pipe")
```
</div>

<div style = "margin-bottom:50px;">
```{r bio239 line 25, include = TRUE, echo = TRUE}
data3 <- read.csv("bio239_spring2022b.csv")
data3
```
</div>

<div style = "margin-bottom:50px;">
```{r bio239 line 32, include = TRUE, echo = TRUE}
data3 <- as_tibble(data3)
data3
```
</div>

<div style = "margin-bottom:50px;">

The starting model contains, as its predictors, the main effect of the response
factor (in our study, the species of insect at a site, or mound) and the main effects
and and mutual interactions of th explanatory factors. The response variable in this model
is the counts. We set up the null model (with a Poisson distribution and an implicit log 
link function) as follows:

</div>

<div style = "margin-bottom:50px;">
```{r bio239 line 49, include = TRUE, echo = TRUE}
glm.x0 <- glm( count ~ species + season * mound, data = data3, family = poisson)
```
</div>

<div style = "margin-bottom:50px;">
We compare this model with alternative models, each extended by one of the two
interactions between the response factor (species) and an explanatory factor:
</div>

<div style = "margin-bottom:50px;">
```{r bio239 line 60, include = TRUE, echo = TRUE}
add1 ( glm.x0, scope = ~species * season * mound, test = "Chisq")
```
</div>

<div style = "margin-bottom:50px;">
We use the `scope` parameter to specify the target model with a full interaction among all 
three factors for the `add1` function, but the function has its own rules regarding the 
sequence of testing interactions of different orders. Consequently, the test of the second-order
interaction (i.e., the "species:season:mound" term) is omitted here, as well as the "season:mound"
interaction term, which is already present in the null model.

Both tested interactions have a significant effect (p < 0.001) and explain a relatively large 
fraction of the variability that was not explained by the null model.
(See the Deviance column: e.g., species:season term explains (100 x (322.99 - 25.78)/322.99) = XX.XX%) 
of the variation unexplained by the glm.x0 model.)

The tests are based on X^2 (Chi-Square) statistics (due to a specific test parameter in the `add1` function call)
but in the `add1` output this statistic is labelled LRT (likelihood ratio test), which it indeed represents.

We also not that the `add1` function scores the 3 models under comparison with the 
information-theoretic statistic (AIC) value and this comparison  confirms the appropriateness of adding both 
interactions to the model. The effect of the mound type seems to be larger than the season, which is in agreement
with the likelihood ratio test.
</div>

<div style = "margin-bottom:50px;">
So we now extend our model by both first-order interactions and call `add1` again with the extended 
model to test the remaining second-order interactions:
</div>


<div style = "margin-bottom:50px;">
```{r bio239 line 93, include = TRUE, echo = TRUE}
glm.x1 <- update ( glm.x0, . ~ . + species:season + species:mound )

add1( glm.x1, scope = ~ species*season*mound, test = "Chisq")
```
</div>

<div style = "margin-bottom:50px;">
This interaction term, however, is not significant (and the AIC value even increases with addition),
so we conclude that the season and the mound type within the neighborhood affect the presence of 
grasshopper species and their count independently.

To provide students with of this research project with a more specific ecological interpretation, we must
work out the direction of the effects we have just revealed. For this, we start from the estimated values 
of our regression coefficients. These coefficients can be obtained with extracting function `coef`. 
</div>

<div style = "margin-bottom:50px;">
```{r bio239 line 111, include = TRUE, echo = TRUE}
coef ( glm.x1 )
```
</div>

<div style = "margin-bottom:50px;">
The number of grasshoppers of different species is significantly affected by the season (Fall vs. Spring)
(X^2,4 = 297.21, p < 0.001) with more grasshoppers in the Fall and the mound location (East vs. North) (X^2,4 = 43.48, p < 0.001), where more grasshoppers on the North mound.

The effects of these two factors on grasshopper abundance and diversity are mutually independent (X^2,4, n.s.) 
</div>




</div>
