---
title: "Statistics: Final Report"
output: html_document
---

```{r bio239 line 6, include = FALSE, echo = FALSE}
require(tidyverse)
require(knitr)
```

<div style = "margin-bottom:50px;">

Download the package `tidyverse`

</div>

<div style = "margin-bottom:50px;">
```{r bio239 line 18, echo = FALSE}
data <- read.csv("csi_bio338_main.csv")
knitr::kable(head(data[, 1:3]), "pipe")
```
</div>

<div style = "margin-bottom:50px;">
```{r bio239 line 25, include = TRUE, echo = TRUE}
data2 <- read.csv("csi_bio338_main.csv")
data2
```
</div>

<div style = "margin-bottom:50px;">
```{r bio239 line 32, include = TRUE, echo = TRUE}
data2 <- as_tibble(data2)
data2
```
</div>

<div style = "margin-bottom:50px;">

The starting model contains, as its predictors, the main effect of the response
factor (in our study, the water fowl species and type of behavior) and the main effects
and and mutual interactions of the explanatory factors. The response variable in this model
is the counts of behavior. We set up the null model (with a Poisson distribution and an implicit log 
link function) as follows:

</div>

<div style = "margin-bottom:50px;">
```{r bio239 line 49, include = TRUE, echo = TRUE}
glm.x0 <- glm( count ~ bird + behavior, data = data2, family = poisson)
```
</div>


<div style = "margin-bottom:50px;">
We compare this model with alternative models, each extended by the interaction between the response 
factor (count) and an explanatory factors of water fowl species and behavior type:
</div>

<div style = "margin-bottom:50px;">
```{r bio239 line 60, include = TRUE, echo = TRUE}
add1 ( glm.x0, scope = ~bird * behavior, test = "Chisq")
```
</div>

<div style = "margin-bottom:50px;">
We use the `scope` parameter to specify the target model with a full interaction among all 
three factors for the `add1` function, but the function has its own rules regarding the 
sequence of testing interactions of different orders. 

The tested interaction has a significant effect (p < 0.001) and explains a relatively large 
fraction of the variability that was not explained by the null model.
(See the Deviance column: e.g., bird:behavior term explains (100 x (162.03 - 0.00)/162.03) = XX.XX%) 
of the variation unexplained by the glm.x0 model.)

The tests are based on X^2 (Chi-Square) statistics (due to a specific test parameter in the `add1` function call)
but in the `add1` output this statistic is labelled LRT (likelihood ratio test), which it indeed represents.
</div>


<div style = "margin-bottom:50px;">
To provide students with of this research project with a more specific ecological interpretation, we must
work out the direction of the effects we have just revealed. For this, we start from the estimated values 
of our regression coefficients. These coefficients can be obtained with extracting function `coef`. 
</div>

<div style = "margin-bottom:50px;">
```{r bio239 line 111, include = TRUE, echo = TRUE}
coef ( glm.x0 )
```
</div>

<div style = "margin-bottom:50px;">
The number of behaviors was significantly affected by the interaction of species of water fowl and 
the behavior type (X^2,3 = 162.03, p < 0.001). 
</div>
